Quantifying passive kelp rafting phenomena and connectivity of nearshore communities around the coast of South Africa through development of a 3D spectral numerical coastal-flow model

### 2020-04-23 thesis_chapter_1.Rmd
house keeping
updated chapter 1

@@

The trajectory of passively drifting objects on the sea surface is influenced by multiple factors, such as water currents, atmospheric wind, wave motion, wave induced currents, gravitational force and buoyancy force. To complicate matters, the previously mentioned factors do not act independently of one another but instead influence one another. Furthermore, the gravitational and buoyancy forces on the object are determined by the objects shape. Therefore, all these factors need to be taken into account when modelling trajectory of drifting objects. Given the local wind, surface current, and the shape and buoyancy of the object is known, it is possible to estimate trajectory by the equation $$V_{drift}=V_{curr} + V_{rel}$$

@@

# Particle dispersion modelling
## Particle dispersion models

The Pol3DD model is able to track virtual particles to simulate water borne dispersion of material. Examples include neutrally buoyant anthropogenic material, larvae, oil spills, outfall discharges and sediment transport [@lebreton2012]. The model uses a second-order accurate advection scheme [@black1990] described as follows: $$\delta\chi=\frac{({u'+ \frac{(u_y\upsilon'-\upsilon_yu')\delta t}{2} )\delta t}}{(1-\frac{u_x\delta t}{2})(1-\frac{\upsilon_y\delta t}{2})-\frac{u_y\upsilon_\chi\delta t^2}{4}}$$

$$\delta y=\frac{({\upsilon' + \frac{(\upsilon_\chi u'-u_\chi\upsilon) \delta t}{2})\delta t}}{(1-\frac{u_y\delta t}{2})(1-\frac{\upsilon_\chi\delta t)}{2}) - \frac{u_\chi\upsilon_y\delta t^2}{4}}$$
where

$$u'= u + \frac{\delta tu_t}{2}$$ $$\upsilon'=\upsilon + \frac{\delta t\upsilon_t}{2}$$
and $u$, $\upsilon$ are orthogonal velocity components, $t$ is time, $\delta t$ is the model time step, $u_\chi$, $u_y$, $\upsilon_y$ are the $u$ and $\upsilon$ spatial velocity gradients, and $u_t$, $\upsilon_t$ are temporal gradients. Horizontal diffusion was modeled as a random walk with separate longitudinal and lateral coefficients set to simulate random turbulence. The distance increments moved by the particle at each time step is calculated as:
$$\Delta X = R_N \sqrt{6E_1\delta t}$$ $$\Delta y = R_N\sqrt{6E_2\delta t}$$
where $R_N$ is a random number in the uniform range (-1,1) and $E_1,E_2$ are the longitudinal and lateral eddy diffusivities, respectively. To estimate the amount of input into the model the authors used a scaled approach to define the release of particles, and where based on previous research performed by @halpern2008.

# Application of particle dispersion models

A study by @lebreton2012 proposed a methodology to track floating debris from source to sink based on descriptions of global waste production and ocean surface currents, using a combination of hydrodynamic and particle dispersion modelling. The particle tracking model applied uses a two stage process, (1) a hydrodynamic model solves the equations of motion to characterise the movement of water within the domain of the model, (2) virtual particles are released into the flow field established in the first step, and particles are allowed to move through hydrodynamic forcing. The authors extracted sea surface currents from the  HYCOM/NCODA ocean circulation modelling system, which is forced by the US Navy's Operational Global Atmospheric Prediction System (NOGAPS). The model also took into account wind stress, wind speed, heat flux, and precipitation and was looped five times to accurately represent ocean circulation patterns for a 30 year period. The authors used the velocity data aquired from HYCOM and coupled it to the Lagrangian particle tracking model Pol3DD which was used to simulate dispersion. No additional wind stress terms were applied to the motion of particles, as this factor had already been incorporated in the HYCOM hydrodynamic data. Potential extra stress for emerged parts of debris was considered negligible as all floating debris are assumed to be fully submerged. The Pol3DD model was also modified so provide additional information on individual particles such as origin, age, and trajectory information of individual particles. Since the model considered a spherical earth, particles were allowed to re-enter the simulation if they passed around the 'globe'. Important to note is only the surface layer was modelled in this study, as much of the evidence identifies surface currents as the main dispersal vector of anthropogenic debris.

@griffin2017 added to the search for flight MH370 by providing estimates of the trajectory of the debris. This was achieved through a combination of an advanced ocean model calibrated with oceanic drifters, and drifters in the form of aircraft parts. Unlike previous work regarding the recovery of MH370 wreckage, this study incorporated windage factors into the trajectory model as well as providing better estimates of Stokes Drift.

@@

Dispersal is recognised as important driver of biodiversity, composition and structure of ecological systems in the marine environment [@bernardes2018;@helmuth1994;@highsmith1985;@macarthur2001;@jackson2010]. These processes are largely dependent on currents for dispersal of migrant populations which ultimately promote connectivity of marine ecosystems [@bernardes2018;@mackas1985;@mackas1985;@zakas2009]. This is particularly true for organisms which lack pelagic larvae, and are reliant on other modes of transport to reach new habitats through passive modes of dispersal from anthropogenic and natural sources. Anthropogenic sources include marine litter and structural debris while examples of organic sources include pumice and macroalgae in the form of kelp-rafts. Kelp-rafts have been identified as an important mode of passive dispersal of marine organisms, such as invertebrates [@helmuth1994; whichmann2012], epiphytes [@], grazers [@nikula2010], invasive species [@lewis2005] and macroalgae themselves [@macaya2005] which either make landfall or are transported offshore where they eventually sink. A kelp-raft consists of an entanglement of macroalgae (one or multiple species) which have been dislodged from the benthic environment through hydrodynamic forces (mostly storms) and are positively buoyant by means of gas-filled pneumatocysts and/or reproductive organs. Kelp-rafts are capable of travelling vast distances [@] and are considered important dispersal vectors in temperate latitudes, such as the Southern California Bight [@hobday2000a] and the Northern Baltic Sea [@].

Kelp-raft abundance has been shown to vary temporally and spatially in the ocean and around coastlines, which is dependent on seasonal growth patterns. For example, studies investigating the dispersal patterns of *Sargassum* in the West Pacific show increased abundance during growth seasons (spring and summer) as more biomass is available to fragment [@deysher1981; @kingsford1992]. Temporal variability may also be related to seasonal storm frequency where more storms in a particular season relate to an increase in kelp-raft abundance. For example a study by @kingsford1995 investigated the contribution of *Macrocystis pyrifera* rafts to habitat complexity in pelagic environments. The authors found that the abundance of *M. pyrifera* rafts increased during seasons where storms occurred more frequently.

The trajectory of kelp rafts are determined by a combination of oceanographic conditions and biological processes. Since kelp rafts float, evidence points largely to wind driven surface currents as the main driver, although the relative importance of wind versus surface currents is largely unknown. For example a study by @harrold1989 used radio-trackers on both natural and artificial *Macrocystis pyrifera* kelp rafts to investigate seasonal trajectory patterns. The results showed that tracked kelp rafts were mostly deposited nearby the source population and that seasonal wind direction was the primary driver of trajectory for kelp rafts submerged roughly ~0.5 - 1m below the surface @harrold1989. However, the authors did note the influence of surface currents for some of their experiments when winds relaxed; which suggests that surface currents may play a larger role in determining trajectory for kelp rafts less exposed to wind or when prevailing winds are low.

>>>>

# Linear wave theory

@@

The trajectory of passively drifting objects on the sea surface is influenced by multiple factors, such as water currents, atmospheric wind, wave motion, wave induced currents, gravitational force and buoyancy force. To complicate matters, the previously mentioned factors do not act independently but instead influence one another. Furthermore, the gravitational and buoyancy forces on the object are determined by the objects shape. Therefore, all these factors need to be taken into account when modelling trajectory of drifting objects. Given the local wind, surface current, and the shape and buoyancy of the object are known, it is possible to estimate trajectory by the equation $$V_{drift}=V_{curr} + V_{rel}$$

@@

@@

Dispersal is recognised as important driver of biodiversity, composition and structure of ecological systems in the marine environment [@bernardes2018;@helmuth1994;@highsmith1985;@macarthur2001;@jackson2010]. These processes are largely dependent on currents for dispersal of migrant populations which ultimately promote connectivity of marine ecosystems [@bernardes2018;@mackas1985;@mackas1985;@zakas2009]. This is particularly true for organisms which lack pelagic larvae, and are reliant on other modes of transport to reach new habitats through passive modes of dispersal from anthropogenic and natural sources. Anthropogenic sources include marine litter and structural debris while examples of organic sources include pumice and macroalgae in the form of kelp-rafts. Kelp-rafts have been identified as an important mode of passive dispersal of marine organisms, such as invertebrates [@helmuth1994; whichmann2012], epiphytes [@macaya2016], grazers [@nikula2010], invasive species [@lewis2005] and macroalgae themselves [@macaya2005] which either make landfall or are transported offshore where they eventually sink. A kelp-raft consists of an entanglement of macroalgae (one or multiple species) which have been dislodged from the benthic environment through hydrodynamic forces (mostly storms) and are positively buoyant by means of gas-filled pneumatocysts and/or reproductive organs. Kelps are capable of travelling vast distances [@Fraser2011] and are considered important dispersal vectors in temperate latitudes, such as the Southern California Bight [@hobday2000a; @hobday2000b].

Kelp-raft abundance has been shown to vary temporally and spatially in the ocean and around coastlines, which is dependent on seasonal growth patterns which ultimatley influences overall biomass of seaweed. For example, studies investigating the dispersal patterns of *Sargassum* in the West Pacific show increased abundance during growth seasons (spring and summer) as more biomass is available to fragment [@deysher1981; @kingsford1992]. Temporal variability may also be related to seasonal storm frequency where more storms in a particular season relate to an increase in kelp-raft abundance. For example a study by @kingsford1995 investigated the contribution of *Macrocystis pyrifera* rafts to habitat complexity in pelagic environments. The authors found that the abundance of *M. pyrifera* rafts increased during seasons where storms occurred more frequently.

The trajectory of kelp rafts are determined by a combination of oceanographic conditions and biological processes. Since kelp  float, evidence points largely to wind driven surface currents as the main driver, although the relative importance of wind versus surface currents is largely unknown. For example a study by @harrold1989 used radio-trackers on both natural and artificial *Macrocystis pyrifera* kelp rafts to investigate seasonal trajectory patterns. The results showed that tracked kelp rafts were mostly deposited nearby the source population and that seasonal wind direction was the primary driver of trajectory for kelp rafts submerged roughly ~0.5 - 1m below the surface @harrold1989. However, the authors did note the influence of surface currents for some of their experiments when winds relaxed; which suggests that surface currents may play a larger role in determining trajectory for kelp rafts less exposed to wind or when prevailing winds are low. Other authors have identified the West Wind Drift as important driver of trajectory of floating *M. pyrifera* [@Fraser2011; ]
